---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data


Before we go and download the data, let's first grab the date: `r Sys.Date()`. Next get the data from the web. The file comes from a repository in my GitHub account that was forked from the original.

```{r download}
source_url <- "https://raw.githubusercontent.com/ASonCourse/RepData_PeerAssessment1/master/activity.zip"

download.file(source_url, destfile = "activity.zip", 
              method = "curl")
```

The downloaded file was compressed, so we will unzip it now:

```{r unzip}
unzip("activity.zip", exdir = ".")
```

It turns out the unzipped file is a comma separated file. We can import that into R easily:

```{r import}
activity <- read.csv("activity.csv", stringsAsFactors = FALSE)
```

The `date` variable in `activity` is of type character and needs to be converted to `date`:

```{r char_to_date}
activity$date <- as.Date(activity$date, format = "%Y-%m-%d")
```

This concludes the loading and preprocessing of the data. [Element #1 required for a complete submission ("Code for reading in the dataset and/or processing the data.")]


## What is mean total number of steps taken per day?

Since we'll be using `group_by()` and related commands to figure this out we need to install and load the `dplyr` package:

```{r install_dplyr}
options(repos="https://cran.rstudio.com" )
if (!require(dplyr)) install.packages("dplyr")
library(dplyr)
```

Calculate mean total number of steps per day:

```{r mean_total_steps}
mean_total_steps <- activity %>%
  group_by(date) %>%
  summarise (daily_steps = sum(steps)) %>%
  summarise(mean_total_steps = mean(daily_steps, na.rm = TRUE))

steps <- round(mean_total_steps, 0)
steps
```

Result: `r steps` steps.


## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
